---
title: "Lab 01 - Functions and Vectors"
author: "Erica Santos"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

    # CODE CHUNKS

        # This is a "code chunk" - R evaluates the code inside and prints results
        # These "chunks" begin with "```{r}" and end with "```"

        # Inside {r}, we can modify the chunk behavior
        # Above, we tell R not to "include" this chunk

    # GLOBAL OPTIONS

        # Typically, the first "chunk" tells R how to create your report
        # Below, we tell R to repeat code and disable messages/warnings

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

    # KNITTING

        # "Knitting" a .RMD file is essentially turning it into a data product
        # Even without your answers, this is ready to knit!

        # Click on "Knit" in the upper-left corner to turn this into a report
        # Notice that this "chunk" will not appear in your knitted report!

```

<br>

# Source Data

The following report analyzes tax parcel data from Syracuse, New York (USA).

View the "Data Dictionary" here: 
[**Syracuse City Tax Parcel Data**](https://watts-college.github.io/cpp-526-fall-2022/labs/syr_parcels.html)

<br>
<br>

# Importing the Data

The following code imports the Syracuse, NY tax parcel data using a URL.

```{r cache = TRUE}

url <- paste0("https://raw.githubusercontent.com/DS4PS/Data",
              "-Science-Class/master/DATA/syr_parcels.csv")

dat <- read.csv(url, 
                strings = FALSE)

```

<br>
<br>

# Previewing the Data

There are several exploratory functions to better understand our new dataset.

We can inspect the first 5 rows of these data using function `head()`.

<br>

```{r}

head(dat, 5)              # Preview a dataset with 'head()'

```

<br>

### Listing All Variables

Functions `names()` or `colnames()` will print all variable names in a dataset.

<br>

```{r}

names(dat)                # List all variables with 'names()'

```

<br>

### Previewing Specific Variables

We can also inspect the values of a variable by extracting it with `$`.

The extracted variable is called a "vector".

<br>

```{r}

head(dat$owner, 10)       # Preview a variable, or "vector"

```

<br>

### Listing Unique Values

Function `unique()` helps us determine what values exist in a variable.

<br>

```{r}

unique(dat$land_use)      # Print all possible values with 'unique()'

```

<br>

### Examining Data Structure

Function `str()` provides an overview of total rows and columns (dimensions), 
variable classes, and a preview of values.

<br>

```{r}

str(object = dat,
    vec.len = 2)          # Examine data structure with 'str()'

```

<br>
<br>

# Questions & Solutions

**Instructions:** Provide the code for each solution in the following "chunks".

*Remember to modify the text to show your answer in human-readable terms.*

<br>

## Question 1: Total Parcels

**Question:** *How many tax parcels are in Syracuse, NY?* 

**Answer:** There are **41502** tax parcels in Syracuse, NY. 

<br>

```{r}

nrow(dat)
```

<br>

## Question 2: Total Acres

**Question:** *How many acres of land are in Syracuse, NY?*

**Answer:** There are **12510.49** acres of land in Syracuse, NY.

<br>

```{r}

sum(dat$acres, na.rm=TRUE)
```

<br>

## Question 3: Vacant Buildings

**Question:** *How many vacant buildings are there in Syracuse, NY?*

**Answer:** There are **1888** vacant buildings in Syracuse, NY.

<br>

```{r}

sum(dat$vacantbuil, na.rm=TRUE)
```


## Question 4: Tax-Exempt Parcels

**Question:** *What proportion of parcels are tax-exempt?*

**Answer:** **10.7** of parcels are tax-exempt.

<br>

```{r}

round(sum(dat$tax.exempt, na.rm = TRUE)/length(dat$tax.exempt)*100,2)
```

<br>

## Question 5: Neighborhoods & Parcels

**Question:** *Which neighborhood contains the most tax parcels?*

**Answer:** **Eastwood** contains the most tax parcels.

<br>

```{r}

sort(table(dat$neighborhood), decreasing= TRUE)

```

<br>

## Question 6: Neighborhoods & Vacant Lots

**Question:** *Which neighborhood contains the most vacant lots?*

**Answer:** **Near Westside** contains the most vacant lots.

<br>

```{r}

tab <- table(dat$neighborhood, dat$land_use)
sort(tab[,"Vacant Land"], decreasing=TRUE)

```

<br>
<br>

