---
output:
  html_document:
    theme: readable
    df_print: kable
    highlight: tango
---

```{r setup, include=FALSE}
# do not change this chunk
knitr::opts_chunk$set(echo=T, eval=T, fig.width=6, fig.height=4, warning=F, message=F )
```


# M1 Lab 1 Submission
**Erica Santos**

# Overview

In this assignment, you will analyze the **USArrests** dataset, which contains statistics on violent crime rates in the United States for each of the 50 states in 1973. You will apply the R skills you learned in this module, including importing data, summarizing information, creating visualizations, and using tidyverse functions.


# Load Libraries

For this assignment, and most others in this class you will need to use the `tidyverse` and `GGally` libraries. Before loading your libraries, if you have not already installed those libraries on the version of R you are currently using, please install them now. You can do that by typing the code below in your console or by using the Tools dropdown menu. You will not need to do this again for future assignments as long as you are using the same computer.
```{r, echo=T,eval=F, message=FALSE, warning=FALSE}
install.packages("tidyverse")
install.packages("GGally")
```

Load the necessary libraries for data analysis and visualization.

```{r}
library(tidyverse)
library(GGally)  # For pairwise scatter plots
```


# Data
The **USArrests** dataset contains the following variables:
- `Murder`: Murder arrests (per 100,000 residents)
- `Assault`: Assault arrests (per 100,000 residents)
- `UrbanPop`: Percent of the population living in urban areas

```{r}
# Read in dataset
us_arrests <- read_csv("Data/M1-US-Arrests-Data.csv")
```

Use the following commands to explore the dataset content:

```{r}
# View first few rows of the dataset
head(us_arrests)

# Get a quick glimpse of the dataset
glimpse(us_arrests)

```



# Question 1

Use the following tasks to practice subsetting data using `filter()` to select specific rows and `select()` to choose specific columns.

## 1.1 
Retrieve all rows where `Murder` is greater than 10 per 100,000 residents.

```{r}
us_arrests %>%
  filter(Murder>10) %>%
  select(State,Murder)

```

## 1.2 
Retrieve all rows where `UrbanPop` is less than 50%.

```{r}
us_arrests %>%
  filter(UrbanPop<50) %>%
  select(State,UrbanPop)

```

## 1.3 
Find the data for the states of **California, Texas, and New York**.

```{r}
us_arrests %>%
  filter(State %in% c("California","Texas","New York"))%>%
  select(State,Murder,UrbanPop)


```

## 1.4 
Select only the `State` and `Murder` columns.

```{r}
us_arrests %>%
  select(State,Murder)


```

## 1.5 
Select the `State`, `Assault`, and `UrbanPop` columns.

```{r}
us_arrests %>%
  select(State,Assault,UrbanPop)
```

## 1.6 
Retrieve only the states where `Assault` is greater than 200, but display only the `State` and `Assault` columns.

```{r}
us_arrests %>%
  filter(Assault>200) %>%
  select(State,Assault)

```

## 1.7 
Retrieve only the states where `Murder` is below 5 and `UrbanPop` is above 70, but display only `State`, `Murder`, and `UrbanPop`.

```{r}
us_arrests %>%
  filter(Murder<5) %>%
  filter(UrbanPop>70) %>%
  select(State,Murder)

```


# Question 2

Create a new variable `high_murder` that equals **1** if `Murder` is greater than the median murder rate and **0** otherwise.

```{r}
us_arrests %>%
  mutate(high_murder = ifelse(Murder > median(Murder), 1, 0))


```


# Question 3

## 3.1 

Create a scatter plot to show the relationship between the number of assault arrests (`Assault`) and murder arrests (`Murder`). 

```{r}
library(ggplot2)
ggplot(us_arrests,aes(x=Assault, y=Murder))+geom_point()

```

## 3.2 

Add a linear trend line to the plot. 

```{r}
library(ggplot2)
ggplot(us_arrests, aes(x=Assault, y=Murder))+geom_point()+geom_smooth(method="lm")

```

## 3.3 

Create histograms for `Murder` and `UrbanPop` to understand their distributions.

```{r}
library(ggplot2)
ggplot(us_arrests, aes(x=Murder))+geom_histogram(binwidth=1)+labs(title="Distribution of Murder Arrest Rates",x="Murder", y="Frequency")+theme_minimal()
```

```{r}
library(ggplot2)
ggplot(us_arrests, aes(x=UrbanPop))+geom_histogram(binwidth=1)+labs(title="Distribution of Urban Population",x="Urban Population (%)", y="Frequency")+theme_minimal()

```

## 3.4

Generate pairwise scatter plots to explore relationships among `Murder`, `Assault`, and `UrbanPop`.

```{r}
library(GGally)
ggpairs(us_arrests[,c("Murder","Assault","UrbanPop")])



```


# Question 4

## 4.1 
Calculate the mean murder rate for states classified as `high_murder` (1) and not high murder (0).

```{r}
library(dplyr)

us_arrests <- us_arrests %>%
  mutate(high_murder = ifelse(Murder > median(Murder), 1, 0))

us_arrests %>%
  group_by(high_murder) %>%
  summarise(mean_murder = mean(Murder))

```

## 4.2 
Create a new variable `high_urban` that equals **1** if `UrbanPop` is greater than the median urban population percentage and **0** otherwise. Then, summarize the mean murder rate by `high_urban`.

```{r}
library(dplyr)

us_arrests <-us_arrests%>%
  mutate(high_urban=ifelse(UrbanPop>median(UrbanPop),1,0))

us_arrests %>%
  group_by(high_urban) %>%
  summarise(mean_murder = mean(Murder))
```

